Time-series modeling of epidemics in complex populations: detecting changes in incidence volatility over time

### 2024-12-20 ms-3/ms.tex
Update manuscript and fig1 code

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 US counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

\subsection*{Data availability and processing}
We applied our method to cumulative COVID-19 case countes in US counties~\citep{nytimes_covid19}.
Population sizes were from The United States Census Bureau~\citep{us_census_county_population_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases. Missing values for new cases were all at the beginning of the pandemic and were imputed as zero. 
Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
We used a width paramter of $\delta = 8$ and set the degrees of freedom parameter for the natural splines to $J = 3$.

>>>>

\begin{abstract}
Trends in infectious disease incidence provide important information about epidemic dynamics and prospects for control. 
Higher-frequency variation around incidence trends can shed light on the processes driving epidemics in complex populations, as transmission heterogeneity, shifting landscapes of susceptibility, and fluctuations in reporting can impact the volatility of observed case counts.
However, measures of temporal volatility in incidence, and how volatility changes over time, are often overlooked in population-level analyses of incidence data, which typically focus on moving averages.
Here we present a statistical framework to quantify temporal changes in incidence dispersion and detect rapid shifts in the dispersion parameter, which may signal new epidemic phases. 
We apply the method to COVID-19 incidence data in 144 United States (US) counties from the January 1st, 2020 to March 23rd, 2023.
Theory predicts that dispersion should be inversely proportional to incidence, however our method reveals pronounced temporal trends in dispersion that are not explained by incidence alone, but which are replicated across counties. 
In particular, dispersion increased around the major surge in cases in 2022, and highly overdispersed patterns became more frequent later in the time series.
These findings suggest that heterogeneity in transmission, susceptibility, and reporting could play important roles in driving large surges and extending epidemic duration. 
The dispersion of incidence time series can contain structured information which enhances predictive understanding of the underlying drivers of transmission, with potential applications as leading indicators for public health response.
\end{abstract}

@@

\subsection*{Data availability and processing}
We applied our method to cumulative COVID-19 case counts in United States (US) counties~\citep{nytimes_covid19}.
Population sizes were from taken the US Census Bureau~\citep{us_census_county_population_2021}.
Cumulative cases for the largest three counties in each state were converted to weekly counts by keeping the last observation from each week and differenced to compute new cases. Missing values for new cases were all at the beginning of the pandemic and were imputed as zero. 
Approximately 0.24\% of new cases were negative due to corrections in the cumulative data. These were also imputed as zero. 
We used a width paramter of $\delta = 8$ week and set the degrees of freedom parameter for the natural splines to $J = 3$.  
[This seems like a wide window and a low-df spline. Why did we choose these and what happens if we use different values? Some sort of sensitivity analysis is probably needed, even if it is just to run the analysis using a smaller window and larger df]

\section*{Results and Discussion}
Our estimates of $\theta$ were robust across the population sizes included in the empirical analysis of U.S. counties. 
However, inaccuracies were more common in counties with small populations outside the examined range.%(\nameref{S1_Fig}). 
We found that the LRT method of testing for sharp changes in $\theta$ is also robust across population sizes represented in the empirical data (Fig\ 1 e, f). 
Adequate test performance is characterized by an average \( p \)-value of 0.5 when the effect size is zero, with decreasing average $p$-values as the effect size increases.

\begin{figure}[!ht]
\includegraphics[width=1\textwidth]{fig1}
\caption{
Detecting changes in dispersion within incidence time series for populations of different sizes. (a, b) Simulated case counts with constant (a) and changing (b) dispersion parameters. (c, d) Dispersion parameters used to generate the data in (a) and (b), respectively. (e, f) LRT performance under varying absolute differences in \(\theta\) (horizontal axis), showing \( p \)-value distributions for population sizes of 50,000 (e) and 10,000,000 (f), representing the range in the empirical data. The null and alternative hypotheses (O and X) correspond to the parameters in (c) and (d). Red vertical lines indicate interquartile ranges.
}\label{fig1}
\end{figure}
